\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-02 papers/jfp/7-experience.tex
Word smithing in Section 7

As a set of authors, we've written a paper about the \Shake build
system~\cite{mitchell2012shake}, applying \Shake to building GHC~\cite{hadrian},
and a paper about the categorisation of build
systems~\cite{mokhov2018buildsystems}. The original motivation of the research
in that final paper was to create ``\Cloud \Shake'', and apply it to the GHC
build system. In this section we tie these papers together, reflecting on our
experience since these papers.

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{\url{https://shakebuild.com}}, and
add a lot of tests. The biggest change in that time period was moving from
blocking threads to continuations for the suspending scheduler. The most visible
change is switching from \hs{*>} to \hs{@\%@>} for defining rules\footnote{A
conflicting \hs{*>} operator was added to the Haskell \hs{Prelude}.}, but almost
all external and internal details remain the same.

We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of the categorisation paper~\cite{mokhov2018buildsystems}. There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. The second
aspect is the power of \Shake tasks, or ``actions'' in the original paper,
which were originally described using this data type:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}

\noindent
This type is equivalent to \hs{Task}~\hs{Monad} -- the two types form an
isomorphism. The original paper uses concrete types \hs{Key} and \hs{Value}, we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}. We define that
\hs{Monad} instance, and use it to write conversion functions to/from our
\hs{Task} type, witnessing the isomorphism:

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

@@

\noindent
Similarly, \Make tasks were described as equivalent to

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Rule k v a = Rule { depends :: [k], action :: [v] -> a }
\end{minted}
\vspace{1mm}

Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, the
data type \hs{Rule}~\hs{k}~\hs{v}~\hs{v} is isomorphic to
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v}, and we can define a similar
\hs{Applicative} instance and conversion functions. By describing these types
using \hs{Task} we are able to describe the differences more concisely
(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
and isn't possible, and explore other constraints beyond just \hs{Monad} and
\hs{Applicative}.

These and other isomorphisms for \emph{second-order functionals}, i.e.
functions of the form

@@

\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
developed to replace a \Make-based build system and solve multiple scalability
and maintainability challenges. As discussed in detail
by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
of two key \Make's shortcomings: (i)~poor abstraction facilities of makefiles,
i.e. the need to program in a single namespace of mutable string variables, and
(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
\Hadrian benefits both from \Shake's features and from the host language
Haskell, making the new GHC build system easier to understand and maintain.
Interestingly, although \Shake is not a self-tracking build
system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
domain-specific language for constructing build command lines, and then tracks
command lines by treating them as a type of values -- an example of
\emph{partial self-tracking} made possible by \Shake's support for value
polymorphism~(\S\ref{sec-polymorphism}).

The development of \Cloud \Shake allows GHC developers to take advantage of
sharing build results between builds, which is particularly promising in the
context of GHC's continuous integration (CI) infrastructure responsible for
building and testing every GHC commit. Building GHC~8.8 from scratch with
default settings takes around 1 hour on Windows using both \Hadrian and the
original \Make-based build system. This includes building the compiler itself,
29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
only a small part of the whole codebase, making the cloud-build functionality
highly desirable. In particular, if a commit does not touch any GHC source files
at all (e.g. only modifies the documentation), \Hadrian can build GHC in under
3~minutes, by simply creating symbolic links to the previously built results
stored in the \Cloud~\Shake cache. Note that a small number of GHC build rules
cannot be cached since they mutate the bundled package database (e.g. to
register a new library); \Cloud~\Shake provides a way to manually label such
build rules, thereby excluding them from caching.

One of the benefits of using \Shake is that we have access to high quality build
profiling information, allowing us to compute critical paths and other metrics,
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
profiling features. This information has shown us that GHC is slow to build
(1~hour on an 4~CPU Windows machine), that more CPUs would not help (on
unlimited CPUs the speed up would be less than 10\%), and that a handful of
build tasks (two anomalously slow Haskell compilations, and some calls to slow
single-threaded \cmd{configure}) are responsible for a significant amount of
build time (at least 10~minutes).

@@

Converting \Shake into \Cloud \Shake was not a difficult process once armed with
a roadmap~\cite{mokhov2018buildsystems}. The key was the introduction of two new
functions:

@@

\noindent
These functions are suspiciously like \hs{recordCT} and \hs{constructCT}
from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
illustrative of the changes required. (We have made some minor changes from
actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
differences.)

\begin{itemize}
\item There are two \hs{Ver} arguments being passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be supplied at once, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Irrelevant differences}: A common problem is that you do not get
      shared caching when you want it. As one example, imagine two users install
      \cmd{gcc} on different paths (say \cmd{/usr/bin} and \cmd{/usr/local/bin})
      -- if these paths are recorded by the build system the users won't share
      cache entries. As another example, consider a compiler that embeds the current time
      in the output -- any users who build that file locally won't get any shared
      caching of subsequent outputs. Possible solutions include using relative paths,
      depending only on version numbers for system binaries (e.g. \cmd{gcc}),
      controlling the environment closely (e.g. using \Nix) and extra flags to
      encourage compilers to be more deterministic.
\item \textbf{Insufficient produced files}: A build rule must declare all files
      it produces, so these can be included in the cache.
      As an example using Haskell, compilation of \cmd{Foo.hs} produces
      \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
      and only ever depend on \cmd{Foo.hi}, but \emph{use} \cmd{Foo.o} after
      depending on \cmd{Foo.hi}, a local build will probably work. However,
      if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
      and subsequent commands may fail (e.g. linking). In practice, most issues encountered
      during the move to cloud builds for GHC were caused by failing to declare produced
      files.
\item \textbf{Insufficient dependencies}: While insufficient dependencies are
      always a problem, the move to a cloud build makes them even more serious.
      If you omit a dependency on a file that changes rarely, each developer will
      have built it locally at least once, and more regularly if they ever do clean
      builds. In contrast, with a cloud build, the file might only be built
      \emph{just once}, and reused forevermore. If things go wrong, a clean build
      is no longer a refuge of last resort.
\end{itemize}

@@

>>>>

We have described a framework that can describe, and indeed execute in
prototype form, a wide spectrum of build systems.  But our ultimate
goal is a practical one: to use these insights to construct a better
build system.  Our earlier work on the \Shake build
system~\cite{mitchell2012shake}, and applying \Shake to
building GHC~\cite{hadrian}, makes progress in that direction.

Based on the thinking in this paper we have extended \Shake to
become \Cloud \Shake, the first cloud-capable build system to support
both early cutoff and monadic dependencies (\S{sec-implementation-cloud}),
and deployed it to GHC's (very substantial) build system, Hadrian~\cite{hadrian}.
In this section we tie these papers together, reflecting on our
experience since these papers.

@@

The original design of \Shake has not changed since the initial paper, but the
implementation has continued to mature -- there have been roughly 5,000
subsequent commits to the \Shake
project\footnote{\url{https://github.com/ndmitchell/shake}}. These commits add
concepts like \emph{resources} (for handling situations when two build tasks
contend on a single external resource), rewriting serialisation to be faster,
documentation including a website\footnote{\url{https://shakebuild.com}}, and
add a lot of tests. The biggest change in that time period was an implementation
change: moving from
blocking threads to continuations for the suspending scheduler.
But otherwise, almost all external and internal details remain the same\footnote{
  The most visible change is purely notational: switching from \hs{*>} to \hs{@\%@>}
  for defining rules, because a conflicting \hs{*>} operator was added to the
  Haskell \hs{Prelude}.}.
We consider the lack of change suggestive that \Shake is based on fundamental
principles -- principles we can now name and describe as a consequence
of this paper.

There are two main
aspects to the original \Shake paper~\cite{mitchell2012shake} that are described
more clearly in this paper. Firstly, the rebuilder can now be described using
verifying step traces~\S\ref{sec-step-traces}, with a much clearer relationship
to the unoptimised verifying traces of~\S\ref{sec-verifying-traces}. Secondly,
in the original paper the tasks (there called ``actions'')
were described in continuation-passing style using this data type\footnote{
The original paper uses concrete types \hs{Key} and \hs{Value}. Here we
generalise these types to \hs{k} and \hs{v}, and also add \hs{a} so that
\hs{Action}~\hs{k}~\hs{v} can be an instance of \hs{Monad}}.
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Action k v a = Finished a
                  | Depends k (v -> Action k v a)
\end{minted}
\vspace{1mm}
\noindent
In this paper we describe tasks more directly, in a monadic (or applicative or functorial)
style. But in fact the two are equivalent: \hs{Task Monad k v} is
isomporphic to \hs{Action k v v}.
To be concrete, the functions \hs{toAction} and \hs{fromAction} witness the
isomorphism in both directions.
\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
instance Monad (Action k v) where
    return = Finished
    Finished x    >>= f = f x
    Depends ds op >>= f = Depends ds $ \v -> op v >>= f

@@

\noindent
Similarly, in the original paper \Make tasks were described as equivalent to

\vspace{1mm}
\begin{minted}[xleftmargin=10pt]{haskell}
data Rule k v a = Rule { depends :: [k], action :: [v] -> a }
\end{minted}
\vspace{1mm}
\noindent
Assuming the lengths of the lists \hs{[@@k]} and \hs{[@@v]} always match, the
data type \hs{Rule}~\hs{k}~\hs{v}~\hs{v} is isomorphic to
\hs{Task}~\hs{Applicative}~\hs{k}~\hs{v}, and we can define a similar
\hs{Applicative} instance and conversion functions.

By describing these types
using \hs{Task} we are able to describe the differences more concisely
(\hs{Monad} vs \hs{Applicative}), use existing literature to determine what is
and isn't possible, and explore other constraints beyond just \hs{Monad} and
\hs{Applicative}.
These and other isomorphisms for \emph{second-order functionals}, i.e.
functions of the form

@@

@@

Converting \Shake into \Cloud \Shake was not a difficult process once armed with
the roadmap in this paper. The key was the introduction of two new
functions:

@@

\noindent
These functions are suspiciously like \hs{recordCT} and \hs{constructCT}
from~\S\ref{sec-constructive-traces}, with their differences perhaps the most
illustrative of the changes required\footnote{
  We have made some minor changes from
actual \Shake, like replacing \hs{Key} for \hs{k}, to reduce irrelevant
differences.}.

\begin{itemize}
\item Two \hs{Ver} arguments are passed to each function. These are
      the versions of the build script, and the rule for this particular key. If
      either version changes then it is as though the key has changed, and
      nothing will match. These versions are important to avoid using stale
      build products from previous versions of the build script.
\item The list of dependencies to \hs{addCloud} is a list of lists, rather than
      a simple list. The reason is that \Shake allows a list of dependencies to
      be specfied simultaneously, so they can all be built in parallel.
\item The \hs{addCloud} function also takes a list of \hs{FilePath}, being the
      files that this rule produces. These produced files include those which
      are output keys from a rule and those declared with the function
      \hs{produces}.
\item The \hs{lookupCloud} function allows an explicit \hs{Nothing} when looking
      up a dependent key, since some keys are not buildable.
\item The \hs{lookupCloud} function returns at most one result, rather than a
      list. This change was made for simplicity.
\end{itemize}

@@

\begin{itemize}
\item \textbf{Irrelevant differences}: A common problem is that you do not get
      shared caching when you want it. As one example, imagine two users install
      \cmd{gcc} on different paths (say \cmd{/usr/bin/gcc} and \cmd{/usr/local/bin/gcc}).
      If these paths are recorded by the build system, the users won't share
      cache entries. As another example, consider a compiler that embeds the current time
      in the output: any users who build that file locally won't get any shared
      caching of subsequent outputs. Possible solutions include using relative paths;
      depending only on version numbers for system binaries (e.g. \cmd{gcc});
      controlling the environment closely (e.g. using \Nix); and extra flags to
      encourage compilers to be more deterministic.
\item \textbf{Insufficient produced files}: A build rule must declare all files
      it produces, so these can be included in the cache.
      As an example using Haskell, compilation of \cmd{Foo.hs} produces
      \cmd{Foo.hi} and \cmd{Foo.o}. If you declare the rule as producing \cmd{Foo.hi},
      and other rules \emph{depend on} \cmd{Foo.hi}, but \emph{also use} \cmd{Foo.o} after
      depending on \cmd{Foo.hi}, a local build will probably work. However,
      if \cmd{Foo.hi} is downloaded from a remote cache, \cmd{Foo.o} will not be present,
      and subsequent commands may fail (e.g. linking). In practice, most issues encountered
      during the move to cloud builds for GHC were caused by failing to declare produced
      files.
\item \textbf{Missing dependencies}: While missing dependencies are
      always a problem, the move to a cloud build makes them even more serious.
      Suppose you omit a dependency on a file \cmd{foo} that is produced by a rule,
      but that changes rarely (e.g. perhaps \cmd{foo} is derived from
      a configuration input file).
      With purely local builds, each developer will build \cmd{foo} at least once (when
      starting from scratch), and more regularly if they ever do clean
      builds. In contrast, with a cloud build, the file might only be built
      \emph{just once}, and reused forevermore. If things go wrong, a clean build
      is no longer a refuge of last resort.
\end{itemize}

@@

\Hadrian is a build system for the Glasgow Haskell Compiler~\cite{ghc}. It was
developed to replace a \Make-based build system and solve multiple scalability
and maintainability challenges. As discussed in detail
by~Mokhov~\etal~\shortcite{hadrian}, most of these challenges were consequences
of two key shortcomings of \Make: (i)~poor abstraction facilities of makefiles,
notably the need to program in a single namespace of mutable string variables, and
(ii)~the lack of dynamic dependencies~(\S\ref{sec-background-shake}).
\Hadrian benefits both from \Shake's features and from the host language
Haskell, making the new GHC build system easier to understand and maintain.

Interestingly, although \Shake is not a self-tracking build
system~(\S\ref{sec-tracking-aspects}), \Hadrian implements a little
domain-specific language for constructing build command lines, and then tracks
command lines by treating them as a type of values -- an example of
\emph{partial self-tracking} made possible by \Shake's support for value
polymorphism~(\S\ref{sec-polymorphism}).

The subsequent development of \Cloud \Shake allows GHC developers to take advantage of
sharing build results between builds, which is particularly promising in the
context of GHC's continuous integration (CI) infrastructure responsible for
building and testing every GHC commit. Building GHC~8.8 from scratch with
default settings takes around 1 hour on Windows using both \Hadrian and the
original \Make-based build system. This includes building the compiler itself,
29~bundled libraries, such as \cmd{base} (each in vanilla and profiled way), and
6~bundled executables, such as \cmd{Haddock}. A single commit typically modifies
only a small part of the whole codebase, making the cloud-build functionality
highly desirable. For example, if a commit does not touch any GHC source files
(e.g. only modifies the documentation), \Hadrian can build GHC from scratch in under
3~minutes, by simply creating symbolic links to the previously build results
stored in the \Cloud~\Shake cache.

A small number of GHC build rules
cannot be cached since they mutate the bundled package database (e.g. to
register a new library); \Cloud~\Shake provides a way to manually label such
build rules, thereby excluding them from caching.

One of the benefits of using \Shake is that we have access to high quality build
profiling information, allowing us to compute critical paths and other metrics;
see Mitchell~\shortcite{mitchell2019ghcrebuildtimes} for an overview of \Shake's
profiling features. This information has shown us, for example, that more CPUs would not help (on
unlimited CPUs the speed up would be less than 10\%), and that a handful of
build tasks (two anomalously slow Haskell compilations, and some calls to slow
single-threaded \cmd{configure}) are responsible for a significant fraction of
build time (at least 15\%).
